Unity in diversity: Children from 17 diverse communities process gaze in similar ways

### 2024-03-14 paper/supplement.Rmd

```{r, echo=FALSE, message=F}
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggpubr)
library(tidyboot)
library(tidybayes)
library(maps)
library(brms)
library(cowplot)
library(ggridges)
library(papaja)
library(coda)
library(ggrepel)
library(knitr)
library(kableExtra)

@@

```{r}
data <- read_csv("../data/gafo-cc-clean-data.csv")%>%
  mutate(continent = ifelse(
    community %in% c("leipzig", "plymouth"), "Europe",
    ifelse(
      community %in% c("akure", "hai||om", "khwe", "chimfunshi", "bandongo", "bayaka", "windhoek", "uganda"), "Africa",
      ifelse(community %in% c("stanford", "mexico", "buenos_aires"), "Americas",
             ifelse(community %in% c("india", "beijing", "turkey"), "Asia", "Oceania"))
    )
  ))%>%
  mutate(continent = factor(continent, levels = c("Americas", "Africa", "Europe", "Asia", "Oceania"), ordered = T))%>%
  mutate(community = recode(community,
                            akure = "Akure (Nigeria)",
                            leipzig = "Leipzig (Germany)",
                            `hai||om` = "Hai||om (Namibia)",
                            khwe = "Khwe (Namibia)",
                            windhoek = "Windhoek (Namibia)",
                            stanford = "Stanford (USA)",
                            chimfunshi = "Chimfunshi (Zambia)",
                            mexico = "Ocuilan (Mexico)",
                            plymouth = "Plymouth (UK)",
                            beijing = "Beijing (China)",
                            india = "Pune (India)",
                            buenos_aires = "Buenos Aires (Argentina)",
                            auckland = "Auckland (New Zealand)",
                            turkey = "Malatya (Türkiye)",
                            bandongo = "Bandongo (Rep. Congo)",
                            bayaka = "BaYaka (Rep. Congo)",
                            uganda = "Nyabyeya (Uganda)"))
```

@@

```{r}
page <- data%>%
  group_by(community,continent)%>%
  distinct(subjid, .keep_all = T)%>%
  mutate(dens = approxfun(density(ageinyears))(ageinyears))%>%
  ggplot(aes(y = community, x = ageinyears))+
  facet_grid(continent~., scales = "free", space = "free_y")+
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0, height = 0),
    point_shape = '|',  point_alpha = 0.5, alpha = 0.7,
  )+
  labs(x = "Age (in years)", y= "")+
  theme_few()+
  theme(strip.background = element_blank(),strip.text.y = element_blank())

pred_agg_i <- data%>%
  mutate(older_children = children - 1 - younger_children)%>%
  group_by(community, subjid)%>%
  summarise(screen = mean(screen, na.rm = T),
            touchscreen = mean(touchscreen, na.rm = T),
            older_children = mean(older_children, na.rm = T),
            household = mean(household, na.rm = T))

pred_agg_g <- data%>%
  mutate(older_children = children - 1 - younger_children)%>%
  group_by(community, continent)%>%
  summarise(screen = mean(screen, na.rm = T),
            touchscreen = mean(touchscreen, na.rm = T),
            older_children = mean(older_children, na.rm = T),
            younger_children = mean(younger_children, na.rm = T),
            children = mean(children, na.rm = T),
            household = mean(household, na.rm = T))%>%
  pivot_longer(names_to = "dem", values_to = "value", cols = c(screen, touchscreen, older_children,children, younger_children, household))

ppred <- pred_agg_g%>%
  filter(dem == "screen" | dem == "touchscreen")%>%
  ggplot(aes(y = community, x = dem, fill = value))+
  geom_point(pch = 22, size = 10, stroke = 0)+
  facet_grid(continent~., scales = "free", space = "free_y")+
  labs(x = "", y= "")+
  scale_fill_viridis_c(name = "Prevalence")+
  theme_few()+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y.right = element_text(angle = 0))
```

@@

@@

@@

@@

@@

Chimfunshi wildlife orphanage is an NGO providing a home for about 150 chimpanzees. The sanctuary is located in the Copperbelt region in the northern part of Zambia close to DRC border. The chimpanzees are housed in eight different enclosures; each of an area up to 121.5Ha. The Sanctuary provides housing, medical care and education for the employees and their families which are in total about 300 people living in 5 different compounds around Chimfunshi.

Chimfunshi itself was founded in 1983 by David and Sheila Siddle who lived in Zambia since the 1950’s. In the past years they rehabilitated chimpanzees who were confiscated from poachers and from obsolete zoos and circuses worldwide. Today the sanctuary is managed by a Board of Trustees to ensure the long-term sustainability.

Throughout the year Zambia has a mainly humid subtropical climate with a dry winter. There are three seasons. The rather cool dry season which lasts from May to August. A short hot dry season which is followed by a hot and humid rainy season from Novmeber to April. Overall, the temperatures are relatively mild due to the altitude of the country in the north but can be cold in winter. The sanctuary is nestled by woodland, savanna and situated near the Kafue river.

The country has about 18 Million inhabitants. The population consists mostly of Bantu-speaking groups whereby Bemba is the largest ethnicity and the most spoken language beside Chewa, Tonga and Lozi. English which is the country's official language, is only spoken by about 2% of the population as a first language. For people with high education levels English is the most commonly used second language. Complex patterns of migrations created a broad linguistic and cultural variety. Due to the big mining industry, the population in the Copperbelt area consists of people from all parts of Zambia and even from neighboring countries.

Zambia is mainly a Christian country (Protestant, Romanic Catholic, Jehovah’s Witnesses and some fundamentalists churches) but due to the Asian community, Hindus and Muslim are represented as well. The northeast and northwest part of Zambia is still strongly influenced by cultural traditions of the Congo basin. Zambian people still value traditional norms within their families, villages and clans. Such norms, ideals and morals are for example  mutual respect and reciprocity.

The main sectors of the economy in Zambia is agriculture and copper mining and smelting especially in the northern Copperbelt region. Despite these large copper deposits, Zambia is one of the poorest countries in the world. Three quarters of the poor live in rural areas. As in other African countries there is a strong trend of movement from rural areas to urban centers whereby men migrate more often and women remain behind in villages. The housing style in rural areas vary a lot by ethnic group. Overall, mud and thatch, brick, or other materials are used for construction.

People working at Chimfunshi wildlife orphanage are living in compounds. These are little settlements made of stone houses with corrugated iron roofs, a well to access clean water, solar systems for electricity and often little gardens to grow crops and/or keeping farm animals (e.g. chickens).

Since independence day Zambia invested a lot in education. By now schooling is compulsory for 7- to 14-year-olds. Primary school lasts for 7 years and the secondary level for 5 years. The Twampane School at Chimfunshi was founded in 2011 to give the children of Chimfunshi employees the opportunity to attend school upto grade 9th. Almost 120 of the students live in the Chimfunshi area, the others come from the surrounding farms and settlements, thus the schools provides education for about 220 students. Children who perform well at school are enabled due to sponsorship programs to attend a secondary school. Furthermore, Chimfunshi offers children in the final grades on-site vocational training and further education. (9)
Depending on the financial possibilities of the families, electrical devices with touch screens such as smartphones or laptops are available. Many still use cellphones and do not own a smartphone, which also makes access to such electronic devices more difficult for some children or even households.

Before a study can be conducted in Chimfunshi, it must be presented to the Scientific Advisory Board of Chimfunshi. The board members and the sanctuary directress deliberate and decide whether a study can be carried out or not. Once these formalities have been clarified, the parents are approached with the help of an on-site assistant. The study program is explained to each family which are interested to participate in one way or the other in a study. Participation is recorded in writing and can be revoked by the parents at any time.  Depending on the study, recruitment takes place in cooperation with the school teachers. (11)

@@

@@

@@

@@

@@

Leipzig is a medium-sized city located in the east of Germany with a population of approximately 620,000. It serves as a significant urban center within the state of Saxony. Geographically, leipzig is situated in the southernmost part of the North German Plain.

@@

* City website for demographic information: https://english.leipzig.de
* Otto, H., & Keller, H. (2016). Is there something like German parenting. Contemporary Parenting: a Global Perspective, 81-94.

@@

>>>>

```{r, echo=FALSE, message=F}
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggpubr)
library(tidyboot)
library(tidybayes)
library(maps)
library(brms)
library(cowplot)
library(ggridges)
library(papaja)
library(coda)
library(ggrepel)
library(knitr)
library(kableExtra)
library(pals)
library(RColorBrewer)

@@

```{r}
data <- read_csv("../data/gafo-cc-clean-data.csv")%>%
  mutate(continent = ifelse(
    community %in% c("leipzig", "plymouth"), "Europe",
    ifelse(
      community %in% c("akure", "hai||om", "khwe", "chimfunshi", "bandongo", "bayaka", "windhoek", "uganda"), "Africa",
      ifelse(community %in% c("stanford", "mexico", "buenos_aires"), "Americas",
             ifelse(community %in% c("india", "beijing", "turkey"), "Asia", "Oceania"))
    )
  ))%>%
  mutate(continent = factor(continent, levels = c("Americas", "Africa", "Europe", "Asia", "Oceania"), ordered = T))%>%
  mutate(community = recode(community,
                            akure = "Akure (Nigeria)",
                            leipzig = "Leipzig (Germany)",
                            `hai||om` = "Hai||om (Namibia)",
                            khwe = "Khwe (Namibia)",
                            windhoek = "Windhoek (Namibia)",
                            stanford = "Stanford (USA)",
                            chimfunshi = "Chimfunshi (Zambia)",
                            mexico = "Ocuilan (Mexico)",
                            plymouth = "Plymouth (UK)",
                            beijing = "Beijing (China)",
                            india = "Pune (India)",
                            buenos_aires = "Buenos Aires (Argentina)",
                            auckland = "Auckland (New Zealand)",
                            turkey = "Malatya (Türkiye)",
                            bandongo = "Bandongo (Rep. Congo)",
                            bayaka = "BaYaka (Rep. Congo)",
                            uganda = "Nyabyeya (Uganda)"))
```
```{r}
qual_col_pals = brewer.pal.info[brewer.pal.info$colorblind == TRUE,]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
colors=kelly(n=n_distinct(data$community)+1)
colors <- colors[ !colors == '#F2F3F4']
```

@@

```{r}
# page <- data%>%
#   group_by(community,continent)%>%
#   distinct(subjid, .keep_all = T)%>%
#   mutate(dens = approxfun(density(ageinyears))(ageinyears))%>%
#   ggplot(aes(y = community, x = ageinyears))+
#   #facet_grid(continent~., scales = "free", space = "free_y")+
#   geom_density_ridges(
#     jittered_points = TRUE,
#     position = position_points_jitter(width = 0, height = 0),
#     point_shape = '|',  point_alpha = 0.5, alpha = 0.7,
#   )+
#   labs(x = "Age (in years)", y= "")+
#   theme_few()+
#   theme(strip.background = element_blank(),strip.text.y = element_blank())
# 
# pred_agg_i <- data%>%
#   mutate(older_children = children - 1 - younger_children)%>%
#   group_by(community, subjid)%>%
#   summarise(screen = mean(screen, na.rm = T),
#             touchscreen = mean(touchscreen, na.rm = T),
#             older_children = mean(older_children, na.rm = T),
#             household = mean(household, na.rm = T))
# 
# pred_agg_g <- data%>%
#   mutate(older_children = children - 1 - younger_children)%>%
#   group_by(community, continent)%>%
#   summarise(screen = mean(screen, na.rm = T),
#             touchscreen = mean(touchscreen, na.rm = T),
#             older_children = mean(older_children, na.rm = T),
#             younger_children = mean(younger_children, na.rm = T),
#             children = mean(children, na.rm = T),
#             household = mean(household, na.rm = T))%>%
#   pivot_longer(names_to = "dem", values_to = "value", cols = c(screen, touchscreen, older_children,children, younger_children, household))
# 
# ppred <- pred_agg_g%>%
#   filter(dem == "screen" | dem == "touchscreen")%>%
#   ggplot(aes(y = community, x = dem, fill = value))+
#   geom_point(pch = 22, size = 10, stroke = 0)+
#   #facet_grid(continent~., scales = "free", space = "free_y")+
#   labs(x = "", y= "")+
#   scale_fill_viridis_c(name = "Prevalence")+
#   theme_few()+
#   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
#   theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.y.right = element_text(angle = 0))
```

@@

Akure is situated in the Ondo State in southwest Nigeria. About 800,000 people live in this medium-sized city, and most of them are from the nearby urban and rural areas.

There are two different seasons in Akure's tropical climate: the wet season, which runs from April to October, and the dry season, which runs from November to March. Rich vegetation, including deep forests and productive agriculture, is what defines the city.

Akure's natural surroundings are enhanced by the abundant vegetation that includes dense forests, beautiful farmlands, and vivid flora, which is fostered by the climate and fertile soils.

Akure boasts a diverse populace, representing a cultural mosaic where the Yoruba, Igbo, and Hausa communities coexist harmoniously. While the Yoruba ethnicity predominates, English serves as a widely spoken language alongside the native tongue.

Over centuries, Akure has cultivated a sense of unity among its residents, transcending ethnic boundaries and nurturing a strong community spirit grounded in mutual respect and cooperation. This communal ethos extends from familial bonds – encompassing both nuclear and extended families – to broader social networks within neighborhoods, workplaces, and religious congregations. Akure's evolution from a traditional Yoruba kingdom to a thriving urban metropolis is deeply embedded in history and has a rich cultural legacy. The landscape is dotted with evidence of old kingdoms and communities, highlighting the city's lasting influence and crucial role in the history of the area.

The primary industry of Akure is agriculture, with a large number of residents working in a variety of farming-related fields including agricultural cultivation, animal husbandry, and the extraction of palm oil. In addition to providing for livelihoods, this traditional subsistence approach maintains cultural customs and connections to the land, tying communities to their agricultural origins.

The variety of architectural styles and types of homes found in the city are reflected in the housing landscape. Akure offers a diverse range of housing options, including modern urban dwellings scattered across the city and ancient mud-brick homes tucked away in rural hamlets. These housing options can accommodate nuclear families, extended kin networks, and communal living arrangements.

A set of fundamental principles that are deeply embedded in the fabric of Akure society define parenting there. Parents are essential in passing on cultural legacy and forming the identities of future generations because they emphasize reverence for elders, community cohesion, and the value of education. Cultural festivals, ceremonies, and rite of passage serve as platforms for imparting values and fostering social cohesion within the community.

The differences in parents' educational attainment in Akure are a reflection of their socioeconomic status and differences in access to formal education. Childcare practices often involve extended family members, particularly grandparents, who play significant roles in nurturing and guiding children.
The city is home to a wide range of educational institutions including public and private schools, that provide extensive curriculum that incorporate standard academic disciplines along with enrichment and cultural education initiatives.

Akure has embraced technology in recent years, with a notable increase in digital adoption, especially among its younger population. The widespread availability of reasonably priced smartphones and other touchscreen devices has contributed to the growing integration of technology into everyday life, affecting both urban and peri-urban areas. This technological revolution has changed social interactions and cultural dynamics within the community by democratizing access to information, communication, and entertainment.

For this study, participants were recruited from churches during Sunday services and residences during community visits within the city of Akure, ensuring a wide representation of its vibrant population.

@@

* ttps://worldpopulationreview.com/world-cities/akure-population

@@

Buenos Aires is the capital of Argentina and located in the North-East of the country, on the shores of the Rio de la Plata. Administratively, the city is autonomous (Autonomous City of Buenos Aires) and has a separate administration from the surrounding Province of Buenos Aires. The city has a population of about 3 million inhabitants and the larger metropolitan area of about 15 million inhabitants. The majority of the population are either of European descent or of mixed descent, and about 2% of the population is of indigenous descent (e.g., Quechua, Guaraní, Mapuche, Aymara). Spanish is the predominant language in Buenos Aires, and the local dialect is known as Rioplatense Spanish.

The city has a humid, subtropical climate and has generally mild weather. The average temperatures range from 11 degrees Celsius (July) to 25 degrees (January), but heat waves are common during the summer months (December to February).

Buenos Aires was first established during Spanish colonization in the 16th century and was involved in trans-Atlantic trade when it became the capital of the Viceroyalty of Rio de la Plata (1776). The War of Independence against the Spanish crown lasted from 1810 to 1816 and was followed by an internal civil war between federalists and centralists over the constitution of the nation state. In the present day, Argentina is a presidential federal republic and Buenos Aires is its political and cultural center.

Moreover, the city is the commercial, financial, and industrial hub of Argentina and a major site for tourism. Buenos Aires has a market economy and the main economic activities center on the service sector (e.g. finance, commerce, tourism, social services) and, to some extent, manufacturing. The city contributes about 20% to the Gross Domestic Product of Argentina. Most people in Buenos Aires live in apartment buildings, but informal settlements have been growing over the last decades. Argentina has experienced one of the highest inflation rates in the world in the past years and poverty has been steadily increasing—though poverty rates in the city of Buenos Aires tend to be lower than in the surrounding Buenos Aires Province.

Commonly, children grow up in nuclear families, but ties with extended family tend to be strong. Generally, family and closeness to family are important cultural values in Argentina. Traditionally, female caregivers had a central role in the family, but this has been undergoing change in recent years, particularly, in middle-class families, with male caregivers taking on caring responsibilities, too. Argentina was the first country in South America to legalize same-sex marriage in 2010 and Buenos Aires in particular is perceived as progressive and liberal city. Some parenting values in middle-class families in Argentina, such as being responsive to the child and providing stimulating encounters, are similar to countries such as the U.S. or France. Caregivers tend to control their children’s schedules/activities, but children do not necessarily perceive this control as negative and caregivers are more permissive with younger children.

Caregivers in the current sample mostly came from a middle-class background and usually had completed tertiary education. In Argentina, children attend nursery/kindergarten from two to three years of age with the last year of kindergarten, from age 5, being compulsory. Children attend primary school from age 6 years and secondary school from age 12 (secondary school lasts 5 or 6 years). Children are schooled in Spanish and usually learn a second language (usually English) from the age of 9. Children in Indigenous communities in other Argentinian Provinces (e.g. Salta Province, Formosa Province) will often speak their respective Indigenous language (e.g. Wichí) as their primary language and schools often rely on bilingual teaching assistants to help during schools lessons (that are predominantly taught in Spanish).
Smartphones, tablets, and computers are common, and children are exposed to these devices from early on.
Children were recruited from one local kindergarten by sending out information letters and consent forms to parents. Children who had parental consent to take part in the study were tested in Spanish in a quiet room in their kindergarten by a local experimenter.

@@

* de Minzi, M. C. R., Lemos, V., Rubilar, J. V. (2014). Argentine culture and parenting styles in Parenting Across Cultures: Childrearing, Motherhood and Fatherhood in Non-Western Cultures, Ed. Selin, H., Springer, 277–292.

* Bornstein, M. H. (1996). Ideas about parenting in Argentina, France, and the United States. International Journal of Behavioural Development, 19, 347–368.

* Buenos Aires Cuidad. Useful information on Buenos Aires. Retrieved from: https://buenosaires.gob.ar/solarcities/en/conoce-buenos-aires

* Benitez, J. (2022). Moving on up? The downside of the high life in Buenos Aires social housing. The Sociological Review. Retrieved from: https://thesociologicalreview.org/magazine/october-2022/verticality/moving-on-up/

* Giménez, J. (2023). La pobreza aumentó al 40.1% en el primer semestre de 2023 y afecta al 56.2% de los niños. Retrieved from: https://chequeado.com/el-explicador/la-pobreza-aumento-al-401-en-el-primer-semestre-de-2023-y-afecta-al-562-de-los-ninos/

* Instituto Nacional de Estadística y Censos, Censo Nacional de Población, Hogares y Viviendas (2022). Retrieved from: https://www.indec.gob.ar/indec/web/Nivel4-Tema-2-41-135

@@

Chimfunshi wildlife orphanage is an NGO providing a home for about 150 chimpanzees. The sanctuary is located in the Copperbelt region in the northern part of Zambia close to the DRC border. The chimpanzees are housed in eight different enclosures; each of an area up to 121.5Ha. The Sanctuary provides education, housing and medical care for the employees and their families which are in total about 300 people living in 5 different compounds around Chimfunshi.

Chimfunshi itself was founded in 1983 by David and Sheila Siddle who lived in Zambia since the 1950’s. In the past years they rehabilitated chimpanzees who were confiscated from poachers and from obsolete zoos and circuses worldwide. To ensure long-term sustainability, the sanctuary is managed by a Board of Trustees.

Zambia experiences a predominantly subtropical climate. The temperatures throughout the year are relatively mild but can be cold in winter due to the altitude of the country in the north. The countries’ climate oscillates between a cool dry season (May to August) followed by a short hot and dry season which turns into a hot and humid rainy season lasting from November to April. The sanctuary is nestled amidst woodland and savanna, close to the Kafue River.

The country, with approximately 18 million inhabitants, consists of a diverse population primarily composed of Bantu-speaking groups. Among these, the Bemba ethnicity stands out as the largest, accompanied by other significant groups such as Chewa, Tonga, and Lozi. While English serves as the official language, it is spoken natively by only around 2% of the population. However, for those with higher education levels, English becomes the predominant second language. The patterns of migrations has contributed to a rich tapestry of linguistic and cultural diversity across the nation., Due to a robust mining industry in the Copperbelt region, the population hailing from various corners of Zambia and even neighboring countries.

Zambia, predominantly a Christian nation, encompasses a diverse religious landscape. Due to the presence of an Asian community, Hindu and Muslim are represented beside the Christian denominations including Protestant, Roman Catholic, Jehovah’s Witnesses, and various fundamentalist churches. Cultural traditions from the Congo basin significantly influence the northeast and northwest regions of Zambia. Despite modernization, Zambian society continues to uphold traditional norms within families, villages, and clans. These enduring values emphasize mutual respect and reciprocity as cornerstones of communal harmony.

Agriculture and copper mining and smelting, particularly in the northern Copperbelt region, are the primary sectors of Zambia’s economy. Despite the abundance of copper deposits, Zambia ranks among the world’s poorest nations. The majority of the impoverished, three-fourths to be precise, reside in rural areas. Similar to other African nations, there’s a pronounced trend of migration from rural to urban areas, with men migrating more frequently while women often stay back in the villages. The architectural style in rural areas significantly differs across ethnic groups, with construction materials ranging from mud and thatch to brick and other resources.

People employed at the Chimfunshi wildlife orphanage reside in compounds. These compounds are small communities comprised of stone houses topped with corrugated iron roofs. They feature wells for clean water access, solar systems for power, and frequently, small gardens for crop cultivation and/or livestock rearing, such as chickens.

Since its Independence Day, Zambia has made significant investments in education. Currently, education is mandatory for children aged 7 to 14. The educational system is structured with a 7-year primary school program followed by a 5-year secondary level. In 2011, the Twampane School was established at Chimfunshi, providing the children of its employees the chance to pursue education up to the 9th grade. Approximately 120 students reside within the Chimfunshi area, while the rest commute from nearby farms and settlements, bringing the total student population to around 220. Sponsorship programs are in place to support academically outstanding students in attending secondary school. Additionally, Chimfunshi provides on-site vocational training and further education for students in the final grades.

The availability of touchscreen electronic devices like smartphones or laptops depends on each family’s financial capacity. While many still utilize cellphones, the lack of smartphone ownership can pose challenges in accessing such devices for some children or households.

Prior to initiating a study at Chimfunshi, it must first be submitted to the Chimfunshi Scientific Advisory Board. The board members, along with the sanctuary directress, review and determine the feasibility of the proposed study. Once the necessary formalities are sorted, an on-site assistant assists in reaching out to the parents. The study program is explained to each family expressing interest in participating in the study. Written consent is obtained for participation, which can be withdrawn by the parents at any point. Depending on the nature of the study, the recruitment process is carried out in collaboration with the school teachers.

@@

The Hai||om are an indigenous group primarily living in the areas in and around the Etosha Pan of northern Namibia. In the current research, participants were recruited in small communities of 150 to 300 people in the Oshikoto and Otjozondjupa regions.

They live in a dry Savannah of northern Namibia, characterized by the arid Namib Desert and the fringes of the Kalahari. The climate is hot and dry, with average temperatures exceeding 30°C and unsteady and minimal rainfall during summer. Vegetation is sparse, consisting mainly of drought-resistant shrubs and grasses with bloom of edible and non-edible plants during rain season.

The Hai||om (i.e., treesleepers), also referred to as Heikum or Hai//om, speak Hai||om, which belongs to the Khoekhoe language family often recognized for containing distinct consonants (i.e., “click” sounds). Local dialects (i.e., =Ākhoe) differ across communities. They traditionally lived in small, semi-nomadic bands, loosely organized around kinship ties and shared leadership roles. Like many other forager communities of southern Africa, the Hai||om value egalitarianism, autonomy, and sharing on demand as foundational schemas organizing their social structure.

The Hai||om are the indigenous people of the Etosha pan, have been living in this area for centuries. During the colonial era and German and South-African occupation of their lands, the Hai||om faced forced relocation from Etosha to the neighboring farmlands. The dispossession of their traditional lands together with political and societal discrimination continue having lasting impact on their communities and livelihoods.

Traditionally, the Hai||om have been hunter-gatherers, relying on their vast knowledge of the desert environment to find food and water. They utilized a variety of techniques for hunting and gathering, including tracking, digging, and setting traps. They exhibited a semi-nomadic lifestyle, taking advantage of seasonal  resource availability. This self-sufficient lifestyle has been impacted by factors like climate change and government programs, but many Hai||om continue to practice traditional subsistence activities whenever possible. Today, the rural Hai||om communities participating in the current research practice a mixed subsistence combining seasonal and occasional farm work, small-scale gardening and animal husbandry, foraging, but also rely on government food programs.

Most of the Hai||om families participating in this research house in small huts constructed from branches, corrugated iron, and animal dung. These dwellings are easily portable and household mobility remains high in these communities. Household composition is flexible, with extended families often living together and individuals moving between and within  communities based on kinship ties and resource availability.

Hai||om child-rearing emphasizes autonomy, egalitarianism, cooperation, and sharing on demand as foundational schemas. Children are actively involved in daily activities from a young age, learning essential skills primarily through observation and participation, rather than direct pedagogy. Parents also instill cultural knowledge and values through storytelling and traditional practices.

Due to their historical marginalization and remote location, access to formal education remains challenging for many Hai||om communities. In one community, children have access to a primary school from age 6 onward, whereas in the other community, children need to travel about 35km to access the next school. In both communities, dropout rates are reported low by teachers, likely due to multiple reasons, such as cultural mismatches between often authoritarian teaching styles and foreign learning content at schools, children’s autonomy in deciding on their learning environments, and secondary schools being located far away from their home communities. Traditional knowledge transfer remains vital, with elders playing a significant role in childcare and education. At the same time, children spend considerable amounts of their daily lives in mixed-age peer groups, which present important learning contexts for these children. Children learn through practical experience and participation in daily activities alongside adults and peers.

Exposure to technology, particularly smartphones and other touchscreen devices, is highly limited among the Hai||om due to their remote location and traditional lifestyle. Electricity is available through solar panels at schools, farm houses, and few households, but the importance of electricity in people’s lives remains minimal. Some individuals, particularly younger generations, may have some exposure to technology through interactions with outsiders or limited access to shared devices, such as cellphones.

It is important to note that this depiction provides an overview on the communities participating in the current research and does not reflect the diversity of livelihoods among urban-dwelling Hai||om people more generally. Their experiences and adaptations vary depending on individual circumstances and ongoing socio-economic changes.

Children were recruited from their homes as part of a long term research project taking place in these communities. They were later tested in quiet locations either at the school building or vacant spaces in their communities. Testing was done by male adult experimenters and a research team known to the communities for about 8 years. The research was approved by the Max Planck Research Ethics Council, the Namibian Commission on Research Science and Technology (Permit Number RPIV01112021), the Namibian San Council, and the Ministry of Education, Arts and Culture of the Republic of Namibia. The study also adhered to guidelines for best practice in cross-cultural research (Bruno et al., 2022). Children received small amounts of candy as a compensation for their participation.

@@

* Bruno, D., Pope-Caldwell, S. M., Haberl, K., Hanus, D., Haun, D., Leisterer-Peoples, S., Mauritz, S., Neldner, K., Sibilsky, A., & Stengelin, R. (2022). Ethical guidelines for good practice in cross-cultural research. Leipzig: Department of Comparative Cultural Psychology, Max Planck Institute for evolutionary anthropology.

* Dieckmann, U. (2012). Born in Etosha: Living and learning in the wild. Legal Assistance Centre.

* Dieckmann, U., Thiem, M., Dirkx, E., Hays, J., Legal Assistance Centre (Namibia), & Desert Research Foundation of Namibia (Eds.). (2014). Scraping the pot: San in Namibia two decades after independence. Land, Environment and Development Project of the Legal Assistance Centre and Desert Research Foundation of Namibia.

* Friederich, R., & Lempp, H. (2014). Etosha: Hai||om heartland: ancient hunter-gatherers and their environment. Namibia Publishing House (Pty) Ltd.

* Widlok, T. (1999). Living on Mangetti: “Bushman” Autonomy and Namibian Independence. Oxford University Press.

@@

The Khwe (also referred to as Khoe, Khwedam, or Barakhoena) people reside in their ancestral lands within today’s Bwabwata National Park in northwestern Namibia and the bordering regions in Angola and Botswana. There is also a Khwe community in today’s South Africa that has moved there while Namibia became independent in 1990. The current research was conducted in the Namibian Zambezi region which is home to numerous small Khwe communities. There are an estimated 3,000 to 5,000 Khwe residing in Bwabwata as of today.

Due to its sub-tropical climate, inhabitants of Bwabwata National Park experience hot and humid summers, whereas other seasons are much more characterized by a hot and dry climate. Bwabwata is a dual-use National Park in which wild animals, including large predators, elephants, and other animals native to the Zambezi region roam freely around the Khwe and Hambukushu settlements in the Park.

The Khwe people speak Khwedam, which is part of the larger Khoe-Kwadi language spectrum. They traditionally lived in small, semi-nomadic bands with leadership based on a combination of age and experience. Today, people live in rural communities of varying size, ranging from one-household settlements to settlements of up to 600 people. Larger settlements have access to schools, medical centers, police stations, and small shops.

The Khwe of Bwabwata are the indigenous people of the region and have resided in the area for centuries. They faced similar historical injustices as other indigenous groups of Southern Africa, such as the Hai||om, resulting from colonization and political marginalization. This includes, but is not limited to, colonial dispossession, forced relocation, and the neglect of their traditional leadership systems. These continue to have lasting impacts on their community and land rights until today.

Traditionally, the Khwe lived as hunter-gatherers, relying on their extensive knowledge of the environment to hunt game, gather bushfood, and find water. This self-sufficient way of life has been impacted by various factors, including the migration of Hambukushu (Bantu) communities to the area, German colonization and South African occupation, governmental resettlement programs, and other environmental changes. Today, Khwe people are no more allowed to leave their settlements freely as part of the anti-poaching regulations of the dual-use National Park. However, traditional subsistence activities continue being an integral part of the cultural identity of many Khwe people. Their traditional subsistence is today supplemented with wage labor, seasonal harvesting programs, small-scale gardening and animal husbandry, as well as governmental and non-governmental support.

Household composition is flexible, with extended families often residing together and individuals moving between groups based on kinship ties and resource availability. Before Namibian independence, the South African Military Forces had multiple camps in the region due to its proximity to the Angolan border. After independence, Khwe communities increasingly settled in the residential infrastructure they had left behind. Other communities live in small huts made of wood, dung, corrugated iron, and other building materials.

Similar to the Hai||om, Khwe child-rearing emphasizes autonomy, egalitarianism as well as strong norms on cooperation and demand sharing. Children actively participate in daily activities from a young age, learning essential hunting, gathering, and survival skills through observation and participation. Parents also transmit cultural knowledge and values through storytelling and traditional practices.

Access to formal education has been historically limited for the Khwe of Bwabwata due to their remote location and marginalization. As of today, formalized education presents a two-edged sword as it is perceived by many as an important opportunity for social improvement, while also contributing to the erosion of cultural and environmental knowledge and skills. In the community sampled here, children do have access to a local kindergarten and a combined school. Many learners attending school did not experience kindergarten education as they grew up in the smaller communities surrounding this larger settlement. Especially in early childhood, traditional knowledge transfer and parenting practices remain crucial, with elders playing a significant role in childcare and education. Children learn through practical experience and participation in daily activities alongside adults. Additionally, recent initiatives aim to improve access to formal education within the national park.

Exposure to technology, particularly smartphones and other touchscreen devices, is generally limited among the Khwe of Bwabwata also due to their remote location, economic hardships, and lack of access to electricity. However, similar to the Hai||om, the level of access can vary depending on proximity to towns and settlements. Some individuals, especially younger generations, may have some exposure through interactions with external communities or shared devices.

It is important to note that this is only an approximate depiction of the situation faced by the communities participating in the current research. It does by no means give credit to the diversity of livelihoods among the Khwe people more generally. Their experiences and adaptations vary depending on individual circumstances and ongoing socio-economic changes.

Children were recruited for a long-term research project after consulting with community elders and representatives. The project was conducted in local schools or kindergartens and received necessary approvals from relevant ethical bodies in Namibia and Germany: The Max Planck Research Ethics Council, the Namibian Commission on Research Science and Technology (Permit Number RPIV01112021), the Namibian San Council, and the Ministry of Education, Arts and Culture of the Republic of Namibia. Self-commitment regarding research ethics aligned with departmental guidelines for best practice in cross-cultural research (Bruno et al., 2022). Testing was carried out by a male adult from the community alongside a familiar research team. The study adhered to strict ethical guidelines for cross-cultural research, and children received small tokens of appreciation for their participation.

@@

* Dieckmann, U., Thiem, M., Dirkx, E., Hays, J., Legal Assistance Centre (Namibia), & Desert Research Foundation of Namibia (Eds.). (2014). Scraping the pot: San in Namibia two decades after independence. Land, Environment and Development Project of the Legal Assistance Centre and Desert Research Foundation of Namibia.

* Nkengbeza, D., Kangumu, B., & Sibuku, C. (2021). The Perceptions of the KHWE (San) Community at Bwabwata on the Importance and Influence of Formal Education (SSRN Scholarly Paper 3886443). https://papers.ssrn.com/abstract=3886443

* Koot, S., Beek, W. V., & Diemer, J. (2016). The Khwe of Namibia. Foragers between Game, Tourism, and Politics. Anthropos, 111(2), 497–512. https://doi.org/10.5771/0257-9774-2016-2-497

* Paksi, A. (2019). ‘We live in a modern time’: : Local perceptions of traditional knowledge and formal education among a Namibian San community. Hunter Gatherer Research, 5(1–2), 119–151. https://doi.org/10.3828/hgr.2019.6

* Paksi, A. (2020). ‘Surviving Development’: Rural development interventions, protected area management, and formal education with the Khwe San in Bwabwata National Park, Namibia. Faculty of Social Sciences, University of Helsinki.

@@

Leipzig is a medium-sized city located in the east of Germany with a population of approximately 620,000. It serves as a significant urban center within the state of Saxony. Geographically, Leipzig is situated in the southernmost part of the North German Plain.

@@

* City website for demographic information: https://english.Leipzig.de
* Otto, H., & Keller, H. (2016). Is there something like German parenting. Contemporary Parenting: a Global Perspective, 81-94.

@@

Plymouth is a coastal city in the South West of England with a population of more than 260,000 inhabitants. About 91% of the population was born in the UK and of those born outside of the UK, the largest proportion comes from the EU (4.5%). 95% of residents identified as White, 2.3% as Asian, and 1% as Black. English is the predominant language spoken in the city.

The city has a temperate oceanic climate with average temperatures ranging from 4 to 20 degrees Celsius. Winters are generally mild, and temperatures rarely drop below zero degrees. Rain is frequent and highest in autumn and winter.
